---
title: "Reproducible Research - Activity Monitoring"
author: "aniceaux"
date: "February 8, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##Loading and processing the data
```{r}
#load library
library(ggplot2)
#READ IN THE ACTIVITY DATASET
activity <- read.csv(file="activity.csv", header = TRUE)
```

##What is mean total of steps taken per day?
```{r}
#Calculate the total number of steps taken per day (ignore the missing values)
totalSteps <- aggregate(steps~date, activity, sum, na.rm = TRUE)
#Calculate mean and median of the total number of steps taken per day (ignore the missing values)
meanSteps <- mean(totalSteps$steps, na.rm = TRUE)
medianSteps <- median(totalSteps$steps, na.rm = TRUE)
```

Total Number of Steps Taken Per Day
```{r, echo=TRUE}
hist(totalSteps$steps, main = "Total Number of Steps Taken Per Day", xlab="Total Steps")
```

##What is the average daily activity pattern?
```{r}
#Make a time series plot of the 5-min interval and avg number of steps, avg'd across all days
fiveMinInterval <- aggregate(steps~interval, activity, mean, na.rm=TRUE)
#Which 5-min interval, on avg across all days contains the max. number of steps?
maxInterval <- fiveMinInterval[which.max(fiveMinInterval$steps),] #ANSWER: Interval 835
```

Avergae Daily Activity Pattern
```{r, echo=TRUE}
plot(x = fiveMinInterval$interval, y=fiveMinInterval$steps, type = "l", main = "Avergae Daily Activity Pattern", xlab = "5-Minute Interval", ylab = "Average Number of Steps")
```

##Imputing missing values
```{r}
#Calculate the total number of missing values (rows with NAs)
totalNAs <- sum(is.na(activity$steps)) #ANSWER: 2304 NAs
#Create new dataset that is equal to the original dataset but w/ the missing data filled in
activityNoNAs <- transform(activity, 
                           steps = ifelse(is.na(activity$steps), 
                                          fiveMinInterval$steps[match(activity$interval,
                                                                      fiveMinInterval$interval)],
                                          activity$steps))
#Make a histogram of the total number of steps take each day; calculate the mean and median number of steps each day
totalStepsNoNAs <- aggregate(steps~date, activityNoNAs, sum)
#Do the values differ form the previous estimates? What is the impact of imputing missing data on the estimates?
meanTotStepsNoNAs <- mean(totalStepsNoNAs$steps, na.rm = TRUE)
medianTotStepsNoNAs <- median(totalStepsNoNAs$steps, na.rm = TRUE)
diffMean <- meanTotStepsNoNAs - meanSteps #there is no difference between the mean
diffMedian <- medianTotStepsNoNAs - medianSteps #there is a diffence of about 1.189
totalDiff <- sum(totalStepsNoNAs$steps) - sum(totalSteps$steps) #there is a difference of about 86129
```

Imput Number of Steps Taken Per Day
```{r, echo=TRUE}
hist(totalStepsNoNAs$steps, main = "Imput Number of Steps Taken Per Day", xlab = "Steps")
```

##Are the differencecs in activity patterns between weekdays and weekends?
```{r}
#Determine weekday or weekend
typeOfDay <- function(date) {
  day <- weekdays(date)
  if (day %in% c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday'))
    return ('weekeday')
  else if (day %in% c('Saturday', 'Sunday'))
    return ('weekend')
  else
    stop ('Invalid')
}
#Create a new factor variable with two levels - "weekday" and "weekend"
activityNoNAs$date <- as.Date(activityNoNAs$date)
activityNoNAs$day <- sapply(activityNoNAs$date, typeOfDay)
#Calculate the total number of steps taken per 5-min interval and avg number of steps taken, avg'd acrpss all weekday and weekends
stepsByDay <- aggregate(steps~interval+day, activityNoNAs, mean)
```

Avergae Daily Activity Pattern in 5-Minute Interval
```{r, echo=TRUE}
ggplot(stepsByDay, aes(interval, steps)) + geom_line() + facet_grid(day~.) + ggtitle("Avergae Daily Activity Pattern in 5-Minute Interval") + labs(x="5-Minute Interval", y="Average Number of Steps")
```